A Large-scale Study of Polymorphism in R

### 2019-03-29 paper/main.tex
going through, consolidating, dealing with comments. at 5.1

\newcommand{\PACKAGES}{11,463\xspace}
\newcommand{\PROGRAMMERS}{?\xspace}
\newcommand{\PERCENTCRAN}{83\%\xspace}
\newcommand{\CRANTOTAL}{13,841\xspace}
\newcommand{\RLOC}{15,050,267\xspace}
\newcommand{\CLOC}{9,373,542\xspace}
\newcommand{\YEARS}{?\xspace}
\newcommand{\INDEXCOINCIDENCE}{3,499\xspace}
\newcommand{\TOTALINDEXY}{6,561\xspace}
\newcommand{\INDEXYPERC}{53\%\xspace}		% TODO Not an accurate count.

@@

We focus on packages hosted on the \emph{Comprehensive R Archive Network} or
CRAN.  With over 13,000 packages, CRAN is the largest repository of software
written in R. It is experiencing sustained growth with an average of size
new packages a day~\cite{LIgges2017}.  Unlike sites like GitHub, CRAN is a
curated collection of packages: A package is only accepted to CRAN if it
abides by a number of well-formedness rules.  Most relevant for our
purposes, packages must have data, examples, vignettes and tests, all of
which must successfully run. From our perspective this means that each
package in CRAN comes with several executable scripts that exercise some of
its functionality.  Notable exceptions to this rule are packages only
containing data, which have no runnable code but are referenced by other
packages \AT{I should collect stats on this}.

Our corpus is a subset of CRAN, consisting of \PACKAGES packages, accounting
for some \PERCENTCRAN of all packages.  These packages have a total of \RLOC
lines of R code and \CLOC lines of C code. For each package, we extracted
all executable code snippets from documentation, vignettes and tests and ran
them independently recording all calls to R functions.  It is noteworthy that
in order to run the scripts in one package, it is often necessary to load a
number of other packages.  In~\cite{issta18}, the authors estimated code
coverage to be around 68\% when including reverse dependencies.  As our infrastructure
adds overhead to script execution, coupled with the fact that some scripts take 
inordinate amounts of time to run, we limit our analysis of any given package to one hour.
We capture and record function signatures on-the-fly, though, so we do collect (partial) data even 
for these long-running packages. \AT{We expect to extend our coverage of CRAN for the next version of this paper.}

@@

\AT{TODO Change "recorded" to "observed" in recordsbypkg}

@@

In R all arguments are passed as promises, unused arguments will be
unevaluated, for these we can not report a type. THUS WE DO WHAT? SAY MORE.

@@

\AT{Do we want to talk about how sometimes we can't eval an argument, even if a function executes successfully?
Like pipe from magrittr.}

We will need a notion of polymorphism when undertaking this analysis. R
is unityped, and the only notion of polymorphism in the language is its use
of dynamic dispatch on the class of an argument.  For our purposes, we will
define 3 simple types of polymorphism, closely connected with the reflection
functions we use.  We start by defining what it means for an {\it argument}
to be polymorphic.  \AT{Could be valuable to have an example here.}

\begin{itemize}
\item an argument is {\it polymorphic in type} if it has been called with at
  least two distinct ``types'' according to the \code{typeof} function and
  our extra analysis.
\item similarly, an argument is {\it polymorphic in class} if it has been
  called with at least two distinct classes according to the \code{class}
  function.

@@

There is an important distinction to be made between the result of
\code{typeof} and that of \code{class} and \code{attributes}: In R, the
\code{typeof} a value cannot change, and this is not the case with class and
attributes.  As mentioned in \AT{an earlier section}, the class of values
can be easily redefined by assigning to the value's class attribute.
Similarly, attributes can be dynamically added, removed, or changed.

This begs the question: what sort of classes and attribute patterns occur
naturally, before possible user tampering?  In terms of classes, the
discussion can be found in Section~\ref{sec:rlang}.  As for attributes:

@@

\item matrices have at least a {\tt dim} attribute with matrix dimensions,
  and optionally a {\tt dimnames} attribute for dimension names.

>>>>

\newcommand{\PACKAGES}{11,463\xspace}
\newcommand{\PROGRAMMERS}{?\xspace}
\newcommand{\PERCENTCRAN}{83\%\xspace}
\newcommand{\CRANTOTAL}{13,841\xspace}
\newcommand{\RLOC}{15,050,267\xspace}
\newcommand{\CLOC}{9,373,542\xspace}
\newcommand{\YEARS}{?\xspace}
\newcommand{\INDEXCOINCIDENCE}{3,499\xspace}
\newcommand{\TOTALINDEXY}{6,561\xspace}
\newcommand{\INDEXYPERC}{53\%\xspace}		% TODO Not an accurate count.
\newcommand{\DATAPKGS}{206\xspace}
\newcommand{\DATAPKGSPERC}{1.5\%\xspace}

@@

We focus on packages hosted on the \emph{Comprehensive R Archive Network} or
CRAN.  With over 13,000 packages, CRAN is the largest repository of software
written in R. It is experiencing sustained growth with an average of size
new packages a day~\cite{LIgges2017}.  Unlike sites like GitHub, CRAN is a
curated collection of packages: A package is only accepted to CRAN if it
abides by a number of well-formedness rules.  Most relevant for our
purposes, packages must have data, examples, vignettes and tests, all of
which must successfully run. From our perspective this means that each
package in CRAN comes with several executable scripts that exercise some of
its functionality.  Notable exceptions to this rule are packages only
containing data, which have no runnable code but are referenced by other
packages.
In our experience, only \DATAPKGS packages had no executable test or example code,
accounting for \DATAPKGSPERC of CRAN.

Our corpus is a subset of CRAN, consisting of \PACKAGES packages, accounting
for some \PERCENTCRAN of all packages.  These packages have a total of \RLOC
lines of R code and \CLOC lines of C code. For each package, we extracted
all executable code snippets from documentation, vignettes and tests and ran
them independently recording all calls to R functions.  It is noteworthy that
in order to run the scripts in one package, it is often necessary to load a
number of other packages.  In~\cite{issta18}, the authors estimated code
coverage to be around 68\% when including reverse dependencies.  As our infrastructure
adds overhead to script execution, coupled with the fact that some scripts take 
inordinate amounts of time to run, we limit our analysis of any given package to one hour.
We capture and record function signatures on-the-fly, though, so we do collect (partial) data even 
for these long-running packages.

@@

\AT{TODO Touch up graphs.}

@@

In R all arguments are passed as promises, unused arguments will be
unevaluated, for these we can not report a type. 
\AT{Unused arguments are still picked up by genthat, so we evaluate them.
I suppose this might lead to issues in some rare circumstances.}

\AT{Do we care about this discussion? This is probably a bit too specific to the implementation.}
Packages which implement metaprogramming mechanisms are also a bit challenging.
For example, consider the {\tt magrittr} package, which implements the Tidyverse's pipe functions.
Refer to Figure~\ref{fig:magpipeex} for a translation of a call to {\tt magrittr}'s pipe ({\tt \%>\%}) function.
\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
mul_by <- function(x, y) { x * y }

4 %>% mul_by(3) === `%>%`(lhs = 4, rhs = mul_by(3)) === mul_by(4, 3)
\end{lstlisting}}\caption{Example and rough translation of \%$>$\% call.}\label{fig:realex}\end{figure}
Our tracer evaluates function arguments, and here \code{rhs = mul_by(3)}, but that's erroneous code (you can't call {\tt mul\_by} without supplying two arguments).
In this situation, our tracer will still register the entire call, but flag the {\tt rhs} argument to indicate to later stages of the analysis that there was an issue evaluating that argument.

@@

We will need a notion of polymorphism, and thus a notion of type, when undertaking this analysis. 
R is unityped, and the only notion of polymorphism in the language is its use of dynamic dispatch on the class of an argument.  
For our purposes, we will define a notion in type roughly in line with R's \code{typeof} reflection function, in some cases augmented by information obtained by \code{class} and \code{attributes} reflection.
We start by defining what it means for an {\it argument} to be polymorphic.

\begin{itemize}
  \item an argument is {\it polymorphic in type} if it has been called with at
  least two distinct types according to our definition of type.

   \item an argument is {\it polymorphic in class} if it has been called with at least two distinct classes according to the \code{class} function.

@@

The main ingredient of our types is the {\tt typeof} a value, and there is an important distinction to be made between the result of \code{typeof} and that of \code{class} and \code{attributes}.
In R, the \code{typeof} a value cannot change, and this is not the case with class and attributes.  
As mentioned in \AT{an earlier section}, the class of values can be easily redefined by assigning to the value's class attribute, and similarly attributes can be dynamically added, removed, or changed.

While we do consider some class and attribute information when constructing our types, it is also valuable to consider them entirely separate from types.
Users are free to define their own classes, and arbitrarily add and remove attributes, and interesting patterns might arise in considering them.
That said, not attribute patterns and classes are necessarily interesting, and indeed some of R's built-in data structures use them to some extend.
For one, data frames have a {\tt data.frame} class (explicitly on the class attribute).
Other standard patterns are:

@@

\item matrices have at least a {\tt dim} attribute with matrix dimensions,
  and optionally a {\tt dimnames} attribute for dimension names. Curiously, the {\tt dim} attribute is all that is needed to define a matrix (i.e., a list with a {\tt dim} attribute {\it is} a matrix).
